An assessment of the biodiversity - ecosystem function relationship in southern African woodlands

### 2020-07-02 manuscript/regional_befr_manuscript.tex
Draft

\textbf{Aim:} Positive correlations between tree \rchange{biodiversity}{species diversity} and ecosystem function have been widely documented, but the nature of the relationship \rchange{in highly disturbed and ecophysiologically stressful systems is less clear}{in southern African savanna/woodlands, which experience high levels of disturbance through fire and ecophysiological stress, is less clear}. \rnew{In disturbed systems it is posited that high levels of disturbance may allow the presence of poor competitors, weakening the correlation between biodiversity and niche complementarity which drives ecosystem function.} Here, we explore the relationship between tree species diversity and aboveground biomass across southern African savannas and woodlands, \rchange{with respect to}{while controlling for} gradients in stem density, resource availability\rnew{, disturbance through fire and herbivory}, and across vegetation types to build a general understanding of the biodiversity - ecosystem function relationship in this understudied ecological context.

@@

\textbf{Methods:} We used a network of \nplots{} \rchange{savanna and woodland}{savanna/woodland tree} plots located across the southern African sub-continent \rold{in which each tree >10 cm diameter was measured and identified to species level}. We used Structural Equation Modelling with path analysis to determine the relationship between tree species diversity and aboveground woody biomass, while accounting for the interactive effects of resource availability \rnew{, disturbance by fire}, and stem density.

\textbf{Results:} \rnew{We found} a positive effect of tree species diversity on aboveground biomass \rold{was demonstrated}, observed \rchange{largely as an}{mainly via the} of increasing woodland structural diversity. We also found that the effect of tree species diversity on biomass increases with stem density. Finally, we found that resource availability affects biomass in southern African woodlands \rchange{largely}{mainly} indirectly, via its effect on species diversity.

@@

% Studies in natural and experimental settings have shown a positive relationship between tree species diversity and ecosystem functionality but many ecosystems are understudied in this regard. In this study we conducted the first regional estimation of the relationship between tree species diversity and above-ground biomass in southern African woodlands, using a network of \nplots{} woodland plots. We used structural equation modelling (SEM) to determine the relationship between tree species diversity and aboveground biomass, with comparison to the effects of resource availability and along a gradient of stem density. A positive effect of tree species diversity on biomass was demonstrated, with most of this as an indirect effect via woodland structural diversity. We found that the effect of tree species diversity on biomass increases with stem density. Finally, we found that resource availability determines biomass in southern African woodlands largely indirectly via its effect on species diversity.

@@

Scientific interest in the relationship between species diversity and ecosystem function springs from both an interest in the factors which structure ecological communities \citep{}, and a more applied interest in determining the effect of global biodiversity loss on ecosystem form and function \citep{}. Numerous studies have shown relationships between biodiversity and ecosystem function (e.g. \citealt{Liang2016, Hooper2012, Cardinale2009}). The strength and direction of these observed Biodiversity-Ecosystem Function Relationships (BEFRs) varies depending on the ecosystem being studied, the ecosystem function(s) of interest \citep{Hector2007}, and the inclusion of environmental covariates in statistical models \citep{Vila2005}, but there appears to be a generalisable positive correlation between biodiversity and ecosystem function \citep{Liang2016}. Over the past decade, many observational studies of the BEFR have been conducted, mostly in wet tropical and temperate forests, and grasslands \citep{Chen2011}. These studies support early findings from small-scale experimental studies \rnew{conducted} predominantly in grassland patches, which began in earnest during the 1990s as concern grew over the global loss of biodiversity \citep{Tilman1994, Tilman2014}.

@@

Compared to \rchange{other forested}{forest} ecosystems, dry tropical woodlands and savannas are highly structured by disturbance, mainly through fire \rchange{and in Africa notably by herbivory also}{and herbivory, with African savannas possessing large herbivores absent from other savannas} \citep{Sankaran2008, Levick2009}. \rnew{Disturbance via human activities such as timber extraction and charcoal processing is also common in African woodlands, often causing high levels of disturbance in localised areas \citep{}}. High levels of disturbance by fire may weaken the role of competition in determinging local species distribution and allow weak competitors to co-exist where they would normally be excluded \citep{Grime1979, Grace1990}. This means that interspecific competition and therefore the effect of niche complementarity, which contributes the majority of the observed biodiversity effect on ecosystem function in temperate and wet tropical forests \citep{Wright2017, Poorter2015, Sande2017a}, may not be as \rchange{apparent}{important} in dry woodland/savanna ecosystems. Instead, stress tolerance and the functional contribution of more abundant species (selection effects) may be the predominant forces influencing ecosystem function \citep{Lasky2014, Tobner2016}. Similarly, more diverse species assemblages may lead to facilitation effects between certain species combinations \rchange{in environments which are more hostile to growth}{under limiting environmental conditions such as low water availability \citep{} or high maximum temperature \citep{}}. Across European forests \citet{Ratcliffe2017} found stronger positive relationships between tree species richness and various ecosystem functions in more arid environments. They suggest that in water-limited ecosystems, facilitative effects and selection effects may be more important than niche complementarity in driving the relationship between species diversity and ecosystem function, \rnew{as competition diminishes in ecosystems where environmental stress limits individual species reaching the bounds of their fundamental niche space, thus reducing the competition which drives niche complementarity effects.} This potential mismatch in the contribution of different mechanisms to the BEFR between dry tropical woodlands and other forested ecosystems demands further investigation \rchange{in order to characterise}{if we are to derive} a generalisable BEFR.

The representation of dry tropical ecosystems in the BEFR literature is poor compared to other ecosystems. \citet{Clarke2017} conducted a meta-analysis of 182 published BEFR studies, finding that only 13\% were conducted in the tropics generally, with 42\% of those being conducted in the wet tropical forests of Costa Rica, which hold many endemic species and unique ecosystem assemblages \citep{Barthlott2005}. A severe lack of study in dry tropical ecosystems, especially given the potential mismatch in BEFR mechanisms described above, suggests that a focus on these ecosystems could greatly strengthen our understanding of a general BEFR and its environmental determinants. Savannas and woodlands are the dominant vegetation type across the southern African region, spanning >4 million km\textsuperscript{2} \citep{White1987, Ratnam2011, Ryan2016} (\autoref{clust_map}). The carbon stored in this vegetation is comparable to that found in the wet forests of the Congo basin, and is of global importance to the carbon cycle \citep{Houghton2009, Mayaux2008}. Climatic conditions and biogeography vary across southern African vegetation, resulting in a diverse range of savanna and woodland tree species assemblages\rchange{, which}{.These} retain the common features of an open tree canopy and an understorey generally dominated by C4 \rchange{grass species}{grasses}. Southern African savannas and woodlands (SAWs) are highly diverse, thought to harbour \textapprox{}8500 plant species of which \rold{there are} >300 \rnew{are} \rchange{tree species}{trees} \citep{Frost1996}, and have been identified by previous studies as a priority for conservation efforts \citep{Byers2001, Mittermeier2003}. Many conservation projects in the region currently aim to conserve biodiversity and woody biomass stocks simultaneously under the directive of the United Nations REDD+ programme or the similar Forest Carbon Partnership Facility (FCPF) \citep{Hinsley2015}. Despite these efforts however, human actions are driving rapid changes in biodiversity, with largely unquantified consequences for ecosystem structure and function.

A small number of studies in SAWs, \rnew{all of which were restricted in the spatial scope to a small region of miombo woodland}, have found that above-ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}. The results of these fine scale studies concur with similar studies in other biomes \citep{}. Studies of the BEFR often find that at fine scales, biodiversity shows a strong effect on ecosystem function, but at broad scales biodiversity effects pale in significance compared to abiotic factors such as climate \citep{Pasari2013}. Due to the highly variable environmental conditions within which SAWs occur \citep{Frost1996}, \rnew{with wide variation in precipitation, diurnal and annual temperature range} and given the potential importance of environment and biogeography in defining the strength and form of a relationship between biodiversity and above ground woody biomass \citep{}, it is important to sample across geographic and environmental gradients to gain understanding of the spatial variation in the relationship between biodiversity and biomass.

@@

\citet{Solbrig1996} writes that SAWs possess structurally diverse tree canopies, with trees occupying distinct layers of the canopy at different growth stages and among species. This structural diversity may be one mechanism through which tree species diversity influences woody biomass. \citet{Kunz2019} found that crown complementarity and crown plasticity both increased with species richness in a seasonally dry subtropical forest. They also found that trees growing in species-rich neighbourhoods exhibited enhanced biomass production. \rchange{Occupation}{Occupancy} of multiple canopy layers allows a \rchange{more full}{fuller} canopy with \rold{a} greater total foliage density, enhancing productivity and allowing greater standing woody biomass in a smaller area via a form of niche complementarity. This mechanism however, which has been supported by experiments and observational studies in temperate and wet tropical ecosystems \citep{Hardiman2011, Stark2012}, may not be relevant in savannas. Instead, the overriding importance of disturbance history may negate the effects of tree species diversity on structural diversity \citep{Grime2012}.

High levels of disturbance in SAWs may moderate the observable BEFR through its effect on ecosystem composition. Fire disturbance in forests has been linked to abundance-dependent mortality among smaller tree stems \citep{Roques2001, Staver2009, Bond2005}. Some species in the regional species pool may be excluded from woodland plots with high levels of disturbance if they are unable to escape the fire bottleneck and grow to become a large tree. Selection effects may therefore be more important in maximising ecosystem function in disturbance prone woodlands. If the regional species pool contains a large number of species, it is more likely that one of them will possess the necessary growth strategy to grow to a large tree with high biomass under an intense disturbance regime.

In this study, we made the first known regional estimation of the biodiversity-ecosystem function relationship across southern African savannas and woodlands (SAWs), using inventory plots which span environmental and biogeographical gradients (\autoref{clust_map}). We used aboveground woody biomass of trees \rnew{as our metric of ecosystem function}, and compared the relative effects of tree species diversity with that of environmental factors known to affect ecosystem productivity and biomass accumulation, namely water availability, energy input and soil fertility. We also investigated the potential moderating effects of environmental covariates on the relationship between tree species diversity and biomass. We incorporated vegetation type (via clustering of plot-level tree species composition), as a factor in our analyses to understand how tree species composition as well as diversity affected ecosystem function and to assess the generality of our results. We used Structural Equation Modelling (SEM) \rchange{and}{with} path analysis as a preferred method to simultaneously account for environmental and biotic factors, \rchange{which may interact their effect}{which may have interacting effects} on ecosystem structure and therefore biomass. Initially, we \rchange{made}{posited} three hypotheses: (1) water availability and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity, (2) \rchange{the effect size}{the strength of the effect} of tree species diversity on woody biomass will increase with plot level stem stocking density, due to an increased importance of niche complementarity as competition increases, and (3) tree species diversity will increase tree structural diversity \rnew{(i.e. physiognomic diversity)}, which will provide an indirect path by which tree diversity increases woody biomass.

@@

The study used \nplots{} woodland monitoring plots from the larger SEOSAW network \citep{seosaw_web} located across 10 countries within southern Africa in the \rchange{so-called miombo woodland eco-region}{miombo ecoregion} (\autoref{clust_map}, \citealt{White1987}). The study \rchange{region spans}{area spacs the core climate space of the region}, with a precipitation gradient from \textapprox{}460 mm y\textsuperscript{-1} in southern Mozambique and southern Zimbabwe to \textapprox{}1700 mm y\textsuperscript{-1} in northern Zambia, Malawi and northern Mozambique. The 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT) of the study sites covers \hullcover{}\% of the pixel-wise climate space of the miombo woodland ecoregion \rold{as defined by} \citep{White1987}, using WorldClim estimates of temperature and precipitation between the year 1970 and 2000 with a pixel size of 30 arc seconds (0.86 km\textsuperscript{2} at the equator) \citep{Fick2017}.

Plots were chosen from a larger pool of 5395 plots based on the quality and completeness of data collection, and plot setup. Plot vegetation was identified under the broad term of ``savanna'', which includes ``woodland'', ``savanna woodland'', and ``tree savanna'', variously defined in other areas of the scientific literature and here referred to collectively as southern African woodlands (SAWs) \citep{Ratnam2011, Hill2010}. Plots with evidence of farming, human resource extraction or experimental treatments such as prescribed burning or herbivore exclusion were excluded from the initial pool. Only plots >0.1 hectares were used in analysis, as area-based biomass estimation from small plots is highly influenced by rare large trees \citep{Stegen2011}, leading to inaccurate estimates. Only plots with a stem density >10 stems ha\textsuperscript{-1} (>10 cm stem diameter) were used, to ensure all plots were within woodland rather than ``grassy savanna'', which are considered a separate biome with very different species composition \citep{Parr2014}.

@@

All stems >10 cm DBH were measured within each plot resulting in a total of \num[group-separator={,}]{\ntrees{}} stems with measurements. A tree may be comprised of multiple stems, but for this analysis each stem is treated as an individual. For each stem, we \rchange{measured}{recorded} species, DBH and tree height to the top of the highest branch material. Height was measured through a variety of means including laser rangefinders, manual clinometers and measuring sticks. When DBH could not be measured at 1.3 m due to trunk abnormalities, it was measured at the closest regular portion of the trunk to 1.3 m. The height of this measurement was recorded and used to estimate the DBH\textsubscript{e} at 1.3 m using a cubic polynomial regression, with parameters estimated using a test dataset from Ryan C., (unpublished), see  \citet{Godlee2020}.

@@

where $\rho$ is the species level mean wood density, $D$ is the DBH\textsubscript{e} at 1.3 m, and $H$ is the tree height. Wood density estimates were taken from the global wood density database for each species where possible \citep{Chave2009, Zanne2009}. Wood density for species without species level estimates was estimated from the means of their respective \rchange{genus}{genera}.

@@

\rnew{We used} Structural Equation Models (SEM) to investigate the determinants of AGB. All SEMs were constructed and analysed in the \verb|lavaan| package \citep{lavaan} in R version 3.6.0 \citep{R2019}. SEM was used because of its suitability for modelling complex causal interactions in ecological systems \citep{Lee2007}. A key aspect to our decision to use SEMs is that they can explicitly model and partition variance \rnew{attributed} to indirect effects, which is challenging in standard multiple regressions. Using SEMs also allowed us to describe latent variables such as ``resource availability'' and ``disturbance'' which have been suggested to act upon biodiversity and biomass/productivity in previous studies despite these factors not having observable measures in our dataset. Structural equation modelling is also necessary to properly account for potential feedback mechanisms between aspects of climate and tree species diversity, which could otherwise increase the chances of Type I error and wrongly attribute inference due to covariance of explanatory variables when using conventional regression analyses \citep{Nachtigall2003}.

Prior to analysis, we specified a conceptual model with factors expected to affect AGB: moisture availability, soil fertility, tree species diversity, tree structural diversity and stem density (\autoref{con_mod}).

\begin{figure}[H]
\centering
	\input{include/path_diagram_concept}
	\caption{Conceptual Directed Acyclic Graph (DAG) showing the theoretical relationships between environmental factors, tree species diversity, tree structural diversity, tree stem density, and AGB. Hypothesised paths of causation are depicted as arrows from predictor to response. Correlations are depicted as curved dotted arrows.}
	\label{con_mod}
\end{figure}

Observed variables were transformed to achieve normality where necessary and standardised to Z-scores prior to analysis (\hyperref[appendixa]{Appendix A}). Standardisation put each latent variable on the same scale, with a mean of zero and a standard deviation of one. Standardisation allows path regression coefficients to be easily compared between paths in the same model to assess their relative effect size, and eliminates confusion in model interpretation arising from the observed variables being on different scales \citep{Beaujean2014}. Standardisation also controls for variables with different orders of magnitude, which could otherwise prevent adequate model estimation from the covariance matrix in \verb|lavaan|. To ensure that observed variables within a latent variable had consistent directions of influence, some observed variables had their sign reversed. For example, soil fertility is expected to decrease as soil sand content increases, so soil percentage sand content was reversed for model fitting. Precipitation seasonality (PS), temperature seasonality (TS), and mean annual temperature (MAT) were also reversed in this way to account for the direction of their effect on moisture availability.

@@

The factor loadings of the observed variable assumed to contribute most to each latent variable were set to 1 as per convention, with other observed variables being allowed to vary \citep{Beaujean2014}.  We tested the robustness of our assumptions with a chi-squared test of all possible combinations of observed variable factor loadings set to 1, while ensuring no factor loadings were in excess of 1. We found no significant difference between model specifications (p>0.05). Full Information Max-Likelihood (FIML) was used in each model to estimate the values of missing data in each latent variable \citep{Cham2017}.

We assessed the role of tree structural diversity and species diversity in determining AGB via a simple mediation model which allowed species diversity to influence AGB both directly and indirectly via structural diversity. To account for variation in stem density, which may covary with species diversity, we also included it as a predictor in our model. To explore variation in the model among woodland vegetation types, we fit the model both at the regional scale and for each vegetation cluster separately. We compared unstandardised path coefficients among these vegetation cluster scale models to understand the effect that vegetation type has on the relationship between tree species diversity, structural diversity, stem density and AGB. Path coefficients show the effect of a path with other paths of inference held constant. Models were estimated using the ``MLM'' estimator, because it is robust to multivariate non-normality \citep{Shapiro1983}. Model fit was evaluated using the robust Comparative Fit Index (CFI), the robust Tucker Lewis Index (TLI), the Root Mean Squared Error \rnew{of Approximation} (RMSEA) and the R\textsuperscript{2} coefficient of determination for AGB. We critically assess model fit in each case, taking into consideration the recommendations of \citet{Hu1999} \rchange{which}{who} define threshold values of acceptability for these model fit indices: CFI >0.85, TLI >0.85, RMSEA <0.15, alongside our judgement of the model estimates.

>>>>

\textbf{Aim:} Positive correlations between tree \rchange{biodiversity}{species diversity} and ecosystem function have been widely documented, but the nature of the relationship \rchange{in highly disturbed and ecophysiologically stressful systems is less clear}{in southern African savanna/woodlands, which experience high levels of disturbance through fire and ecophysiological stress, is less clear}. \rnew{In disturbed systems it is posited that high levels of disturbance may allow the presence of poor competitors, weakening the correlation between biodiversity and niche complementarity which drives ecosystem function.} Here, we explore the relationship between tree species diversity and aboveground biomass across southern African savannas and woodlands, \rchange{with respect to}{while controlling for} gradients in stocking density, resource availability\rnew{, disturbance through fire}, and across vegetation types to build a general understanding of the biodiversity - ecosystem function relationship in this understudied ecological context.

@@

\textbf{Methods:} We used a network of \nplots{} \rchange{savanna and woodland}{savanna/woodland tree} plots located across the southern African sub-continent \rold{in which each tree >10 cm diameter was measured and identified to species level}. We used Structural Equation Modelling with path analysis to determine the relationship between tree species diversity and aboveground woody biomass, while accounting for the interactive effects of resource availability \rnew{, disturbance by fire}, and stocking density.

\textbf{Results:} \rnew{We found} a positive effect of tree species diversity on aboveground biomass \rold{was demonstrated}, observed \rchange{largely as an}{mainly via the} of increasing woodland structural diversity. We also found that the effect of tree species diversity on biomass increases with stocking density. Finally, we found that resource availability affects biomass in southern African woodlands \rchange{largely}{mainly} indirectly, via its effect on species diversity.

@@

% Studies in natural and experimental settings have shown a positive relationship between tree species diversity and ecosystem functionality but many ecosystems are understudied in this regard. In this study we conducted the first regional estimation of the relationship between tree species diversity and above-ground biomass in southern African woodlands, using a network of \nplots{} woodland plots. We used structural equation modelling (SEM) to determine the relationship between tree species diversity and aboveground biomass, with comparison to the effects of resource availability and along a gradient of stocking density. A positive effect of tree species diversity on biomass was demonstrated, with most of this as an indirect effect via woodland structural diversity. We found that the effect of tree species diversity on biomass increases with stocking density. Finally, we found that resource availability determines biomass in southern African woodlands largely indirectly via its effect on species diversity.

@@

In order to understand the effects of global biodiversity change, it is necessary to explore the relationship between biodiversity and ecosystem function \citep{Tilman2014}. Numerous studies have shown relationships between biodiversity and ecosystem function (e.g. \citealt{Liang2016, Hooper2012, Cardinale2009}). The strength and direction of the Biodiversity-Ecosystem Function (BEF) relationship varies depending on the ecosystem being studied, the ecosystem function(s) of interest \citep{Hector2007}, and the inclusion of environmental covariates in statistical models \citep{Vila2005}, but there appears to be a generalisable positive correlation between biodiversity and ecosystem function \citep{Liang2016}. Over the past decade, many observational studies of the BEF relationship have been conducted, mostly in wet tropical and temperate forests, and grasslands \citep{Chen2011}. These studies support early findings from small-scale experimental studies \rnew{conducted} predominantly in grassland patches, which began in earnest during the 1990s as concern grew over the global loss of biodiversity \citep{Tilman1994, Tilman2014}. Despite these concerted efforts, we continue to lack a nuanced, ecosystem agnostic understanding of the complex interactions between biodiversity, environment, and ecosystem function.

@@

Compared to \rchange{other forested}{forest} ecosystems, dry tropical woodlands and savannas are highly structured by disturbance, mainly through fire \rchange{and in Africa notably by herbivory also}{and herbivory, with African savannas possessing large herbivores absent from other savannas} \citep{Sankaran2008, Levick2009}. \rnew{Disturbance via human activities such as timber extraction and charcoal processing is also common in African woodlands, often causing high levels of disturbance in localised areas \citep{}}. High levels of disturbance by fire may weaken the role of competition in determinging local species distribution and allow weak competitors to co-exist where they would normally be excluded \citep{Grime1979, Grace1990}. This means that interspecific competition and therefore the effect of niche complementarity, which contributes the majority of the observed biodiversity effect on ecosystem function in temperate and wet tropical forests \citep{Wright2017, Poorter2015, Sande2017a}, may not be as \rchange{apparent}{important} in dry woodland/savanna ecosystems. Instead, stress tolerance and the functional contribution of more abundant species (selection effects) may be the predominant forces influencing ecosystem function \citep{Lasky2014, Tobner2016}. Similarly, more diverse species assemblages may lead to facilitation effects between certain species combinations \rchange{in environments which are more hostile to growth}{under limiting environmental conditions such as low water availability \citep{} or high maximum temperature \citep{}}. Across European forests \citet{Ratcliffe2017} found stronger positive relationships between tree species richness and various ecosystem functions in more arid environments. They suggest that in water-limited ecosystems, facilitative effects and selection effects may be more important than niche complementarity in driving the relationship between species diversity and ecosystem function, \rnew{as competition diminishes in ecosystems where environmental stress limits individual species reaching the bounds of their fundamental niche space, thus reducing the competition which drives niche complementarity effects.} This potential mismatch in the contribution of different mechanisms to the BEF relationship between dry tropical woodlands and other forested ecosystems demands further investigation \rchange{in order to characterise}{if we are to derive} a generalisable BEF relationship.

The representation of dry tropical ecosystems in the BEF relationship literature is poor compared to other ecosystems. \citet{Clarke2017} conducted a meta-analysis of 182 published BEF relationship studies, finding that only 13\% were conducted in the tropics generally, with 42\% of those being conducted in the wet tropical forests of Costa Rica, which hold many endemic species and unique ecosystem assemblages \citep{Barthlott2005}. A severe lack of study in dry tropical ecosystems, especially given the potential mismatch in BEF relationship mechanisms described above, suggests that a focus on these ecosystems could greatly strengthen our understanding of a general BEF relationship and its environmental determinants. Savannas and woodlands are the dominant vegetation type across the southern African region, spanning >4 million km\textsuperscript{2} \citep{White1987, Ratnam2011, Ryan2016} (\autoref{clust_map}). The carbon stored in this vegetation is comparable to that found in the wet forests of the Congo basin, and is of global importance to the carbon cycle \citep{Houghton2009, Mayaux2008}. Climatic conditions and biogeography vary across southern African vegetation, resulting in a diverse range of savanna and woodland tree species assemblages\rchange{, which}{.These} retain the common features of an open tree canopy and an understorey generally dominated by C4 \rchange{grass species}{grasses}. Southern African savannas and woodlands are highly diverse, thought to harbour \textapprox{}8500 plant species of which \rold{there are} >300 \rnew{are} \rchange{tree species}{trees} \citep{Frost1996}, and have been identified by previous studies as a priority for conservation efforts \citep{Byers2001, Mittermeier2003}. Many conservation projects in the region currently aim to conserve biodiversity and woody biomass stocks simultaneously under the directive of the United Nations REDD+ programme or the similar Forest Carbon Partnership Facility (FCPF) \citep{Hinsley2015}. Despite these efforts however, human actions are driving rapid changes in biodiversity, with largely unquantified consequences for ecosystem structure and function.

A small number of studies in southern African woodlands, \rnew{all of which were restricted in the spatial scope to a small region of miombo woodland}, have found that above-ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}. The results of these fine scale studies concur with similar studies in other biomes \citep{}. Studies of the BEF relationship often find that at fine scales, biodiversity shows a strong effect on ecosystem function, but at broad scales biodiversity effects pale in significance compared to abiotic factors such as climate \citep{Pasari2013}. Due to the highly variable environmental conditions within which southern African woodlands occur \citep{Frost1996}, \rnew{with wide variation in precipitation, diurnal and annual temperature range} and given the potential importance of environment and biogeography in defining the strength and form of a relationship between biodiversity and above ground woody biomass \citep{}, it is important to sample across geographic and environmental gradients to gain understanding of the spatial variation in the relationship between biodiversity and biomass.

@@

\citet{Solbrig1996} writes that southern African woodlands possess structurally diverse tree canopies, with trees occupying distinct layers of the canopy at different growth stages and among species. This structural diversity may be one mechanism through which tree species diversity influences woody biomass. \citet{Kunz2019} found that crown complementarity and crown plasticity both increased with species richness in a seasonally dry subtropical forest. They also found that trees growing in species-rich neighbourhoods exhibited enhanced biomass production. \rchange{Occupation}{Occupancy} of multiple canopy layers allows a \rchange{more full}{fuller} canopy with \rold{a} greater total foliage density, enhancing productivity and allowing greater standing woody biomass in a smaller area via a form of niche complementarity. This mechanism however, which has been supported by experiments and observational studies in temperate and wet tropical ecosystems \citep{Hardiman2011, Stark2012}, may not be relevant in savannas. Instead, the overriding importance of disturbance history may negate the effects of tree species diversity on structural diversity \citep{Grime2012}.

High levels of disturbance in southern African woodlands may moderate the observable BEF relationship through its effect on ecosystem composition. Fire disturbance in forests has been linked to abundance-dependent mortality among smaller trees \citep{Roques2001, Staver2009, Bond2005}. Some species in the regional species pool may be excluded from woodland plots with high levels of disturbance if they are unable to escape the fire bottleneck and grow to become a large tree. Selection effects may therefore be more important in maximising ecosystem function in disturbance prone woodlands. If the regional species pool contains a large number of species, it is more likely that one of them will possess the necessary growth strategy to grow to a large tree with high biomass under an intense disturbance regime.

In this study, we made the first known regional estimation of the biodiversity-ecosystem function relationship across southern African savannas and woodlands, using inventory plots which span environmental and biogeographical gradients (\autoref{clust_map}). We used aboveground woody biomass of trees \rnew{as our metric of ecosystem function}, and compared the relative effects of tree species diversity with that of environmental factors known to affect ecosystem productivity and biomass accumulation, namely water availability, energy input and soil fertility. We also investigated the potential moderating effects of environmental covariates on the relationship between tree species diversity and biomass. We incorporated vegetation type (via clustering of plot-level tree species composition), as a factor in our analyses to understand how tree species composition as well as diversity affected ecosystem function and to assess the generality of our results. We used Structural Equation Modelling (SEM) \rchange{and}{with} path analysis as a preferred method to simultaneously account for environmental and biotic factors, \rchange{which may interact their effect}{which may have interacting effects} on ecosystem structure and therefore biomass. Initially, we \rchange{made}{posited} three hypotheses: (1) water availability and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity, (2) \rchange{the effect size}{the strength of the effect} of tree species diversity on woody biomass will increase with plot level stocking density, due to an increased importance of niche complementarity as competition increases, and (3) tree species diversity will increase tree structural diversity \rnew{(i.e. physiognomic diversity)}, which will provide an indirect path by which tree diversity increases woody biomass.

@@

The study used \nplots{} woodland monitoring plots from the larger SEOSAW network \citep{seosaw_web} located across 10 countries within southern Africa in the \rchange{so-called miombo woodland eco-region}{miombo ecoregion} (\autoref{clust_map}, \citealt{White1987}). The study \rchange{region spans}{area spans the core climate space of the region}, with a precipitation gradient from \textapprox{}460 mm y\textsuperscript{-1} in southern Mozambique and southern Zimbabwe to \textapprox{}1700 mm y\textsuperscript{-1} in northern Zambia, Malawi and northern Mozambique. The 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT) of the study sites covers \hullcover{}\% of the pixel-wise climate space of the miombo woodland ecoregion \rold{as defined by} \citep{White1987}, using WorldClim estimates of temperature and precipitation between the year 1970 and 2000 with a pixel size of 30 arc seconds (0.86 km\textsuperscript{2} at the equator) \citep{Fick2017}.

Plots were chosen from a larger pool of 5395 plots based on the quality and completeness of data collection, and plot setup. Plot vegetation was identified under the broad term of ``savanna'', which includes ``woodland'', ``savanna woodland'', and ``tree savanna'', variously defined in other areas of the scientific literature and here referred to collectively as southern African woodlands \citep{Ratnam2011, Hill2010}. Plots with evidence of farming, human resource extraction or experimental treatments such as prescribed burning or herbivore exclusion were excluded from the initial pool. Only plots >0.1 hectares were used in analysis, as area-based biomass estimation from small plots is highly influenced by rare large trees \citep{Stegen2011}, leading to inaccurate estimates. Only plots with a stocking density >50 trees ha\textsuperscript{-1} (>10 cm stem diameter) were used, to ensure all plots were within woodland rather than ``grassy savanna'', which are considered a separate biome with very different species composition \citep{Parr2014}.

@@

All stems >10 cm DBH were measured within each plot resulting in a total of \num[group-separator={,}]{\ntrees{}} stems with measurements. A tree may be comprised of multiple stems and so tree-level diversity estimates were used to prevent bias towards species which readily coppice. For each tree, we \rchange{measured}{recorded} species, DBH and tree height to the top of the highest branch material. Height was measured through a variety of means including laser rangefinders, manual clinometers and measuring sticks. When DBH could not be measured at 1.3 m due to trunk abnormalities, it was measured at the closest regular portion of the trunk to 1.3 m. The height of this measurement was recorded and used to estimate the DBH\textsubscript{e} at 1.3 m using a cubic polynomial regression, with parameters estimated using a test dataset from Ryan C., (unpublished), see \citet{Godlee2020}.

@@

where $\rho$ is the species level mean wood density, $D$ is the DBH\textsubscript{e} at 1.3 m, and $H$ is the tree height. Wood density estimates were taken from the global wood density database for each species where possible \citep{Chave2009, Zanne2009}. Wood density for species without species level estimates was estimated from the means of their respective \rchange{genus}{genera}. For stems where tree height was unknown, the plot location was used to estimate tree height, according to \citet{Chave2014}.

@@

\rnew{We used} Structural Equation Models (SEM) to investigate the determinants of AGB. All SEMs were constructed and analysed in the \verb|lavaan| package \citep{lavaan} in R version 3.6.0 \citep{R2019}. SEM was used because of its suitability for modelling complex causal interactions in ecological systems \citep{Lee2007}. A key aspect to our decision to use SEMs is that they can explicitly model and partition variance \rnew{attributed} to indirect effects, which is challenging in standard multiple regressions. Using SEMs also allowed us to describe latent variables such as ``resource availability'', ``soil fertility'', and ``disturbance'' which have been suggested to act upon biodiversity and biomass/productivity in previous studies despite these factors not having observable measures in our dataset. SEM is also necessary to properly account for potential feedback mechanisms between aspects of \rchange{climate}{environment} and tree species diversity, which could otherwise increase the chances of Type I error and wrongly attribute inference due to the covariance of explanatory variables when using conventional regression analyses \citep{Nachtigall2003}.

Prior to analysis, we specified a conceptual model with factors expected to affect AGB: moisture availability, soil fertility, disturbance through fire, tree species diversity, tree structural diversity and stocking density (\autoref{con_mod}).

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{concept}
	\caption{Conceptual Directed Acyclic Graph (DAG) showing the theoretical relationships between environmental factors, tree species diversity, tree structural diversity, stocking density, and AGB. Hypothesised paths of causation are depicted as arrows from predictor to response.}
	\label{con_mod}
\end{figure}

Observed variables were transformed to achieve normality where necessary and standardised to Z-scores prior to analysis (\hyperref[appendixa]{Appendix A}). Standardisation put each latent variable on the same scale, with a mean of zero and a standard deviation of one. Standardisation allows path regression coefficients to be easily compared between paths in the same model to assess their relative effect size, and eliminates confusion in model interpretation arising from the observed variables being on different scales \citep{Beaujean2014}. Standardisation also controls for variables with different orders of magnitude, which could otherwise prevent adequate model estimation from the covariance matrix in \verb|lavaan|. To ensure that observed variables within a latent variable had consistent directions of influence, some observed variables had their sign reversed. For example, overall moisture availability is expected to decrease as soil sand content increases, therefore sand content was reversed for use in the moisture availability latent variable. Precipitation seasonality, and temperature stress were also reversed in this way to account for the direction of their effect on moisture availability.

@@

The factor loadings of the observed variable assumed to contribute most to each latent variable were set to one, as per convention, with other observed variables being allowed to vary \citep{Beaujean2014}.  We tested the robustness of our assumptions with a chi-squared test of all possible combinations of observed variable factor loadings set to one, while ensuring no factor loadings were in excess of one. We found no significant difference between model specifications (p>0.05). Full Information Max-Likelihood (FIML) was used in each model to estimate the values of missing data in each latent variable \citep{Cham2017}.

We assessed the role of tree species diversity and tree structural diversity in determining AGB via a simple mediation model which allowed species diversity to influence AGB both directly and indirectly via structural diversity. To account for variation in stocking density, which may covary with species diversity, we also included it as an observed variable in our model. To explore variation in the model among woodland vegetation types, we fit the model both at the regional scale and for each vegetation cluster separately. We compared unstandardised path coefficients among these vegetation cluster scale models to understand the effect that vegetation type has on the relationship between tree species diversity, structural diversity, stocking density and AGB. Path coefficients show the effect of a path with other paths of inference held constant. Models were estimated using the ``MLM'' estimator, because it is robust to multivariate non-normality \citep{Shapiro1983}. Model fit was evaluated using the robust Comparative Fit Index (CFI), the robust Tucker Lewis Index (TLI), the Root Mean Squared Error \rnew{of Approximation} (RMSEA) and the R\textsuperscript{2} coefficient of determination for AGB. We critically assess model fit in each case, taking into consideration the recommendations of \citet{Hu1999} \rchange{which}{who} define threshold values of acceptability for these model fit indices: CFI >0.85, TLI >0.85, RMSEA <0.15, alongside our judgement of the model estimates.

To explore the hypothesis that niche complementarity effects increase in strength as stocking density increases, we repeatedly sub-sampled the available plot dataset to create \subn{} datasets of similar size with varying median stocking density. We used each of these datasets separately to fit the model including only tree species and structural diversity latent variables to predict AGB. We excluded the effect of stocking density on AGB and the correlation between stocking density and species diversity from this model as we deliberately controlled stocking density in our subsampling. We then examined how the unstandardised path coefficients for each path in the SEM varied according to the median stocking density of subsampled datasets.
